Galois Slicing as Differentiation

### 2025-07-11 main/language.tex
Typo.
[ tex ] expand on the examples; emphasise the type dependency

Let the signature
$\Sigma_{\mathsf{num}} = (\{\mathsf{num}\}, \{\mathrm{zero} : \tyUnit \to \mathsf{num}, \mathrm{add} :
\mathsf{num} \times \mathsf{num} \to \mathsf{num}\})$. This signature suffices to write the simple query
function in \exref{introduction-example}, where we interpret the $\mathrm{Label}$ type as the sum
$\tyUnit \tySum \tyUnit$ and the labels $\mathsf{a}$ and $\mathsf{b}$ as $\tmInl{\tmUnit}$ and
$\tmInr{\tmUnit}$. We consider several interpretations of $\Sigma_{\mathsf{num}}$ in $\Fam(\LatGal)$ and their
behaviour on the selection-and-sum query from \exref{introduction-example}.
\begin{enumerate}
\item If we take $\sem{\mathsf{num}}_{\PrimTy} = \Disc(\mathbb{R})$, with $\sem{\mathrm{zero}}_{\PrimOp}$ and
  $\sem{\mathrm{add}}_{\PrimOp}$ the embeddings of the usual zero and addition functions via $\Disc$, then the
  resulting interpretation contains no approximation information. The approximation lattices for all types are
  always the trivial one-point lattice, and the forward and reverse approximation maps are necessarily the
  identity.
\item To get interesting approximation information, we take
  $\sem{\mathsf{num}}_{\PrimTy} = \Lift(\Disc(\mathbb{R}))$, using the lifting monad from
  \secref{models-of-total-gps:lifting}, with $\sem{\mathrm{zero}}_{\PrimOp}$ and
  $\sem{\mathrm{add}}_{\PrimOp}$ defined from the unlifted interpretations above and the monad structure. This
  interpretation recovers the behaviour given in \exref{introduction-example}: running the reverse
  approximation map at the input $\mathsf{a}$ with the top approximation maps indeed reveals that only the
  numbers in the rows tagged with that label in the input are used, and likewise for $\mathsf{b}$.
\item More quantitative approximation information can be obtained by using the interval approximation
  interpretation from \exref{intervals-and-maxima-elements} and \secref{interval-approx}. We let
  $\sem{\mathsf{num}}_{\PrimTy} = \mathbb{R}_{\mathit{intv}}$ and interpret addition using the morphism given
  in \secref{interval-approx}.

  Recall that $\mathrm{query}\,\mathsf{a}\,\mathit{db} = 1$, so in the reverse direction we can pick an
  interval containing $1$ to discover the largest (i.e. least in the order) intervals that will give rise to
  this output as \emph{independent} changes to the input. For example, if we pick
  $[\frac{9}{10},\frac{11}{10}]$ as the interval, then:
  \begin{displaymath}
    (\partial\mathrm{query}\,\mathsf{a}\,\mathsf{db})_r([\frac{9}{10},\frac{11}{10}]) = [(\cdot,[-\frac{1}{10},\frac{1}{10}]),(\cdot,\bot),(\cdot,[\frac{9}{10},\frac{11}{10}])]
  \end{displaymath}
  Indicating that, to achieve an output within $[\frac{9}{10},\frac{11}{10}]$ for this query, either the first
  $\mathsf{a}$ row could be in $[-\frac{1}{10},\frac{1}{10}]$ or the second one could be in
  $[\frac{9}{10},\frac{11}{10}]$.
\end{enumerate}

@@

We can now carry out the systematic insertion of approximation points that we foreshadowed in
\secref{models-of-total-gps:lifting}, using \citet[\S 3.1]{notions-of-computation}'s monadic CBN
translation. We use the $T$ monad from \secref{tagging-monad} because it can be defined in terms of the
language constructed we already have. This requires that we have a signature $\Sigma$ that includes a
primitive type to be interpreted as the approximation object $\mathbb{A}$ and primitive operations to be
interpreted as the monoid operations on this object.

TODO: describe the syntactic CBN translation; show an example on the query example.

>>>>

Let the signature
$\Sigma_{\mathsf{num}} = (\{\mathsf{num}\}, \{\mathrm{zero} : \tyUnit \to \mathsf{num}, \mathrm{add} :
\mathsf{num} \times \mathsf{num} \to \mathsf{num}\})$. This signature suffices to write the simple query
function in \exref{introduction-example}, where we interpret the $\mathrm{Label}$ type as the sum
$\tyUnit \tySum \tyUnit$ and the labels $\mathsf{a}$ and $\mathsf{b}$ as $\tmInl{\tmUnit}$ and
$\tmInr{\tmUnit}$. We consider several interpretations of $\Sigma_{\mathsf{num}}$ in $\Fam(\LatGal)$ and their
behaviour on the selection-and-sum query from \exref{introduction-example}. First, we note the type of the
reverse approximation map in this case. The type of the approximation maps depends on the input value. Our example input database was $\mathit{db} = [(\mathsf{a}, 0), (\mathsf{b}, 1), (\mathsf{a}, 1)]$, meaning that the type of the reverse approximation map for this database and any label $l$ is:
\begin{displaymath}
  \begin{array}{l}
    (\partial \mathrm{query})_r(l, \mathit{db}) : \\
    \quad \partial\sem{\mathsf{num}}(\mathrm{query}\,(l,\mathit{db})) \multimap \mathbb{1} \times (\mathbb{1} \times \partial\sem{\mathsf{num}}(0)) \times (\mathbb{1} \times \partial\sem{\mathsf{num}}(1)) \times (\mathbb{1} \times \partial\sem{\mathsf{num}}(1)) \times \mathbb{1}
  \end{array}
\end{displaymath}
in the category $\JoinSLat$, where $\partial\sem{\mathsf{num}}(x)$ is the lattice of approximations of the
number $x$ determined by our interpretation. In the codomain, the first four $\mathbb{1}$s correspond to the
positions of labels in the input, which we are not approximating, and the final $\mathbb{1}$ is the terminator
of the list. Note how, even if the $\partial\sem{\mathsf{num}}(x)$ does not depend on $x$ the type of the
output is still dependent on the shape of the input list: type dependency is used in a fundamental way in our
interpretation.

\begin{enumerate}
\item If we take $\sem{\mathsf{num}}_{\PrimTy} = \Disc(\mathbb{R})$, with $\sem{\mathrm{zero}}_{\PrimOp}$ and
  $\sem{\mathrm{add}}_{\PrimOp}$ the embeddings of the usual zero and addition functions via $\Disc$, then the
  resulting interpretation contains no approximation information. We have
  $\partial\sem{\mathsf{num}}(x) = \mathbb{1}$ so the type of $(\partial \mathrm{query})_r$ is trivial.
\item We take $\sem{\mathsf{num}}_{\PrimTy} = \Lift(\Disc(\mathbb{R}))$, using the lifting monad from
  \secref{models-of-total-gps:lifting}, with $\sem{\mathrm{zero}}_{\PrimOp}$ and
  $\sem{\mathrm{add}}_{\PrimOp}$ defined from the unlifted interpretations above and the monad structure. The
  type of the reverse map now becomes:
  \begin{displaymath}
    (\partial \mathrm{query})_r(l, \mathit{db}) : \{\top,\bot\} \multimap \mathbb{1} \times (\mathbb{1} \times \{\top,\bot\}) \times (\mathbb{1} \times \{\top,\bot\}) \times (\mathbb{1} \times \{\top,\bot\}) \times \mathbb{1}
  \end{displaymath}
  where every position that corresponds to a number has been tagged with $\top$ for ``present'' and $\bot$ for
  ``not present''. This interpretation recovers the behaviour given in \exref{introduction-example}: running
  the reverse approximation map at the input $\mathsf{a}$ with the top approximation maps indeed reveals that
  only the numbers in the rows tagged with that label in the input are used, and likewise for $\mathsf{b}$.
\item More quantitative approximation information can be obtained by using the interval approximation
  interpretation from \exref{intervals-and-maxima-elements} and \secref{interval-approx}. We let
  $\sem{\mathsf{num}}_{\PrimTy} = \mathbb{R}_{\mathit{intv}}$ and interpret addition using the morphism given
  in \secref{interval-approx}.

  Recall that $\mathrm{query}\,\mathsf{a}\,\mathit{db} = 1$, so in the reverse direction we can pick an
  interval containing $1$ to discover the largest (i.e. least in the order) intervals that will give rise to
  this output as \emph{independent} changes to the input. For example, if we pick
  $[\frac{9}{10},\frac{11}{10}]$ as the interval, then:
  \begin{displaymath}
    (\partial\mathrm{query})_r(\mathsf{a},\mathsf{db})([\frac{9}{10},\frac{11}{10}]) = \cdot, (\cdot,[-\frac{1}{10},\frac{1}{10}]),(\cdot,\bot),(\cdot,[\frac{9}{10},\frac{11}{10}]), \cdot
  \end{displaymath}
  Indicating that, to achieve an output within $[\frac{9}{10},\frac{11}{10}]$ for this query, either the first
  $\mathsf{a}$ row could be in $[-\frac{1}{10},\frac{1}{10}]$ or the second one could be in
  $[\frac{9}{10},\frac{11}{10}]$, and it does not matter what the number in the second entry was.
\end{enumerate}

@@

We can now carry out the systematic insertion of approximation points that we foreshadowed in
\secref{models-of-total-gps:lifting}, using \citet[\S 3.1]{notions-of-computation}'s monadic CBN
translation. We use the $T$ monad from \secref{tagging-monad} because it can be defined in terms of the
language constructs we already have. This requires that we have a signature $\Sigma$ that includes a
primitive type to be interpreted as the approximation object $\mathbb{A}$ and primitive operations to be
interpreted as the monoid operations on this object.

The monadic CBN translation is standard, and entirely determined by the translation on types, so we only
define the type translation $\cbn{-}$ here:
\begin{mathpar}
  \cbn{\rho} = \rho

  \cbn{\sigma \tySum \tau} = T(\cbn{\sigma}) \tySum T(\cbn{\tau})

  \cbn{\tyUnit} = \tyUnit

  \cbn{\sigma \tyProd \tau} = T(\cbn{\sigma}) \tyProd T(\cbn{\tau})

  \cbn{\sigma \tyFun \tau} = T(\cbn{\sigma}) \tyFun T(\cbn{\tau})

  \cbn{\tyList\;\tau} = \tyList\,(T(\cbn{\tau}))
\end{mathpar}
Thus, the CBN translation on types inserts a use of the monad $T(X) = \mathbb{A} \times X$ at the point ``just
underneath'' every type former. In this case, we are describing the type of data annotated at every
level. Note that our lists here are still ``strict'', an alternative approach would be to consider ``lazy''
lists that wrap the tail of every node in a $T$ as well.

We illustrate the effect of the CBN translation on the query example from
\exref{introduction-example}. Applying $\cbn{-}$ to the type of the database
